### 2025-06-28 paper/quarto_aaai/paper.tex
Adjust conclusion, currently at 7 pages of text

As our results indicate, counterfactual training produces models that
are more explainable. Nonetheless, it brings about three important
limitations.\\
\indent \emph{CT increases the training time of models.} CT can be more
time-consuming than conventional training regimes. While higher numbers
of CEs per iteration positively impact the quality of solutions, they
also increase the amount of computations. Relatively small grids with
270 settings can take almost four hours for more demanding datasets on a
high-performance computing cluster with 34 2GB CPUs.\footnote{See
  supplementary appendix for computational details.} Three factors
attenuate this effect: (1) CT amortizes the cost of CEs for the training
samples; (2) it can retain its value when used as a ``fine-tuning''
technique for conventionally-trained models; and (3) it yields itself to
parallel execution, which we have leveraged for our own experiments.\\
\indent \emph{Immutable features may have proxies.} We propose an
approach to protect immutable features and thus increase the
actionability of the generated CEs. However, it requires that model
owners define the mutability constraints for (all) features considered
by the model. Even if all immutable features are protected, there may
exist proxies that are mutable (and hence should not be protected) but
preserve enough information about the principals to hinder the
protections. Delineating actionability is a major open challenge in the
AR literature (see, e.g., \citep{venkatasubramanian2020philosophical})
impacting the capacity of CT to fulfill its intended goal.\\
\indent \emph{Interventions on features may impact fairness.} We provide
a tool that allows practitioners to modify the sensitivity of a model
with respect to certain features, which may have implication for the
fair and equitable treatment of decision subjects. Model owners could
misuse this solution by enforcing explanations based on features that
are more difficult to modify by some (group of) individuals. For
example, consider the \emph{Adult} dataset used in our experiments,
where \emph{workclass} or \emph{education} may be more difficult to
change for underpriviledged groups. When applied irresponsibly, CT could
result in an unfairly assigned burden of recourse
\citep{sharma2020certifai}, threatening the equality of opportunity in
the system \citep{bell2024fairness}. Nonetheless, these phenomena are
not specific to CT.\\
\hfill \break \indent We also highlight several important directions for
future research. Firstly, it is an interesting challenge to extend CT
beyond classification settings. Our formulation relies on the
distinction between non-target class(es) \(y^{-}\) and target class(es)
\(y^{+}\) to generate counterfactuals through Equation~\ref{eq-obj}.
While \(y^{-}\) and \(y^{+}\) can be arbitrarily defined, CT requires
the output space \(\mathcal{Y}\) to be discrete. Thus, it does not apply
to ML tasks where the change in outcome cannot be readily quantified.
Focus on classification models is a common restriction in research on
CEs and AR. Other settings have attracted some interest (e.g.,
regression in \citep{spooner2021counterfactual}), but there is little
consensus how to robustly extend the notion of CEs.\\
\indent Secondly, our approach is susceptible to training instabilities.
This problem has been recognized for JEMs \citep{grathwohl2020your} and
even though we depart from the SGLD-based sampling, we still encounter
considerable variability in the outcomes. CT is exposed to two potential
sources of instabilities: (1) the energy-based contrastive divergence
term in Equation~\ref{eq-div}, and (2) the underlying counterfactual
explainers. We find several promising ways to mitigate this problem:
regularizing energy (\(\lambda_{\text{reg}}\)), generating sufficiently
many counterfactuals during each epoch, and including only mature
counterfactuals for contrastive divergence.\\
\indent Finally, we believe that it is possible to substantially improve
hyperparameter selection procedures. Our method benefits from the tuning
of certain key hyperparameters (see Section~\ref{sec-hyperparameters}).
In this work, we have relied exclusively on grid search for this task.
Future work on CT could benefit from investigating more sophisticated
approaches. Notably, CT is iterative which makes methods such as
Bayesian or gradient-based optimization applicable (see, e.g.,
\citep{bischl2023hyperparameter}). \hfill \break

To conclude, state-of-the-art machine learning models are prone to
learning complex representations that cannot be interpreted by humans
and existing post-hoc explainability approaches cannot guarantee that
the explanations agree with the model's learned representation of data.
As a step towards addressing this challenge, we introduced
counterfactual training, a novel training regime that incentivizes
highly-explainable models. Our approach leads to explanations that are
both plausible---compliant with the underlying data-generating
process---and actionable---compliant with user-specified mutability
constraints---and thus meaningful to their recipients. Through extensive
experiments we demonstrate that CT satisfies its objective while
promoting robustness and preserving the predictive performance of the
models. It can also be used to fine-tune conventionally-trained models
and achieve similar gains. Finally, this work showcases that it is
practical to improve models \emph{and} their explanations at the same
time.

>>>>

As our results indicate, counterfactual training produces models that
are more explainable. Nonetheless, these advantages come at the cost of
two important limitations.

\emph{CT increases the training time of models.} CT can be more
time-consuming than conventional training regimes. While higher numbers
of CEs per iteration positively impact the quality of solutions, they
also increase the amount of computations. In our experiments, grids with
270 settings took up to four hours for more demanding datasets on a
high-performance computing cluster with 34 2GB CPUs.\footnote{See
  supplementary appendix for computational details.} Three factors
attenuate this effect: (1) CT yields itself to parallel execution, which
we have leveraged for our own experiments; (2) it amortizes the cost of
CEs for the training samples; (3) as demonstrated, it retains its value
when used as a ``fine-tuning'' technique for conventionally-trained
models.

\emph{Interventions on features have implications for fairness.} We
provide a tool that allows practitioners to modify the sensitivity of a
model with respect to certain features. Model owners can use our
solution to support the fair and equitable treatment of decision
subjects, but they could also misuse it by enforcing explanations based
on features that are more difficult to modify by some (group of)
individuals. When applied irresponsibly, CT could result in an unfairly
assigned burden of recourse \citep{sharma2020certifai}, threatening the
equality of opportunity in the system \citep{bell2024fairness}.
Additionally, CT requires that model owners define the mutability
constraints for the features considered by the model. Even if all
immutable features are protected, there may exist proxies that are
mutable --- and hence should not be protected --- but preserve
sufficient information about the principals to hinder the protections.
Deciding on actionability is still a major open challenge in the AR
literature \citep{venkatasubramanian2020philosophical} impacting the
capacity of CT to fulfill its intended goal. As such, neither of these
phenomena is ultimately specific to CT.

We also highlight several important directions for future research.
Firstly, it is an interesting challenge to extend CT beyond
classification settings. Our formulation relies on the distinction
between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
and \(y^{+}\) can be arbitrarily defined, CT requires the output space
\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
where the change in outcome cannot be readily discretized. Focus on
classification is a common choice in research on CEs and AR. Other
settings have attracted some interest, e.g., regression
\citep{spooner2021counterfactual}, but there is little consensus how to
robustly extend the notion of CEs.\\
Secondly, CT is susceptible to instabilities in training. This problem
has been recognized for JEMs \citep{grathwohl2020your} and even though
we depart from the SGLD-based sampling, we still encounter variability
in outcomes. CT is exposed to two potential sources of training
instabilities: (1) the energy-based contrastive divergence term in
Equation~\ref{eq-div}, and (2) the underlying counterfactual explainers.
We find several promising ways to mitigate this problem: regularizing
energy (\(\lambda_{\text{reg}}\)), generating sufficiently many
counterfactuals during each epoch, and including only mature
counterfactuals for contrastive divergence.\\
Finally, we believe that it is possible to substantially improve
hyperparameter selection procedures, and thus performance. In this work,
we have relied exclusively on grid searches for this task. Future work
on CT could benefit from investigating more sophisticated approaches.
Notably, our method is iterative which makes methods such as Bayesian or
gradient-based optimization applicable \citep{bischl2023hyperparameter}.

To conclude, state-of-the-art machine learning models are prone to
learning complex representations that cannot be interpreted by humans.
Existing explainability solutions cannot guarantee that explanations
agree with the model's learned representation of data. As a step towards
addressing this challenge, we introduced counterfactual training, a
novel training regime that incentivizes highly-explainable models. Our
approach leads to explanations that are both plausible (compliant with
the underlying data-generating process) and actionable (compliant with
user-specified mutability constraints), and thus meaningful to their
recipients. Through extensive experiments, we demonstrated that CT
satisfies its objective while promoting robustness and preserving the
predictive performance of models. It can also be used to fine-tune
conventionally-trained models and achieve similar gains. Thus, our work
showcases that it is practical to simultaneously improve models and
their explanations.

### 2025-06-28 paper/sections/discussion.qmd
Adjust conclusion, currently at 7 pages of text

As our results indicate, counterfactual training produces models that are more explainable. Nonetheless, it brings about three important limitations.  
`\indent`{=latex} *CT increases the training time of models.* CT can be more time-consuming than conventional training regimes. While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations. Relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect: (1) CT amortizes the cost of CEs for the training samples; (2) it can retain its value when used as a "fine-tuning" technique for conventionally-trained models; and (3) it yields itself to parallel execution, which we have leveraged for our own experiments.  
`\indent`{=latex} *Immutable features may have proxies.* We propose an approach to protect immutable features and thus increase the actionability of the generated CEs. However, it requires that model owners define the mutability constraints for (all) features considered by the model. Even if all immutable features are protected, there may exist proxies that are mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections. Delineating actionability is a major open challenge in the AR literature (see, e.g., [@venkatasubramanian2020philosophical]) impacting the capacity of CT to fulfill its intended goal.  
`\indent`{=latex} *Interventions on features may impact fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects. Model owners could misuse this solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider the *Adult* dataset used in our experiments, where *workclass* or *education* may be more difficult to change for underpriviledged groups. When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness]. Nonetheless, these phenomena are not specific to CT.  
`\hfill \break`{=latex}
`\indent`{=latex} We also highlight several important directions for future research. Firstly, it is an interesting challenge to extend CT beyond classification settings. Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj. While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily quantified. Focus on classification models is a common restriction in research on CEs and AR. Other settings have attracted some interest (e.g., regression in [@spooner2021counterfactual]), but there is little consensus how to robustly extend the notion of CEs.  
`\indent`{=latex} Secondly, our approach is susceptible to training instabilities. This problem has been recognized for JEMs [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter considerable variability in the outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers. We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals for contrastive divergence.  
`\indent`{=latex} Finally, we believe that it is possible to substantially improve hyperparameter selection procedures. Our method benefits from the tuning of certain key hyperparameters (see @sec-hyperparameters). In this work, we have relied exclusively on grid search for this task. Future work on CT could benefit from investigating more sophisticated approaches. Notably, CT is iterative which makes methods such as Bayesian or gradient-based optimization applicable (see, e.g., [@bischl2023hyperparameter]).
`\hfill \break`{=latex}

>>>>

As our results indicate, counterfactual training produces models that are more explainable. Nonetheless, these advantages come at the cost of two important limitations.

*CT increases the training time of models.* CT can be more time-consuming than conventional training regimes. While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations. In our experiments, grids with 270 settings took up to four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect: (1) CT yields itself to parallel execution, which we have leveraged for our own experiments; (2) it amortizes the cost of CEs for the training samples; (3) as demonstrated, it retains its value when used as a "fine-tuning" technique for conventionally-trained models.

*Interventions on features have implications for fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features. Model owners can use our solution to support the fair and equitable treatment of decision subjects, but they could also misuse it by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness]. Additionally, CT requires that model owners define the mutability constraints for the features considered by the model. Even if all immutable features are protected, there may exist proxies that are mutable --- and hence should not be protected --- but preserve sufficient information about the principals to hinder the protections. Deciding on actionability is still a major open challenge in the AR literature [@venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal. As such, neither of these phenomena is ultimately specific to CT.

We also highlight several important directions for future research. Firstly, it is an interesting challenge to extend CT beyond classification settings. Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj. While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily discretized. Focus on classification is a common choice in research on CEs and AR. Other settings have attracted some interest, e.g., regression [@spooner2021counterfactual], but there is little consensus how to robustly extend the notion of CEs.  
Secondly, CT is susceptible to instabilities in training. This problem has been recognized for JEMs [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter variability in outcomes. CT is exposed to two potential sources of training instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers. We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals for contrastive divergence.  
Finally, we believe that it is possible to substantially improve hyperparameter selection procedures, and thus performance. In this work, we have relied exclusively on grid searches for this task. Future work on CT could benefit from investigating more sophisticated approaches. Notably, our method is iterative which makes methods such as Bayesian or gradient-based optimization applicable [@bischl2023hyperparameter].
